---
title: Senate Coalitions
subtitle: An Agent-Based Model
author: 
  - name: 
      given: Max H. 
      family: Feinleib
    orcid: 0009-0002-9604-3533
    email: maxfeinleib2024@u.northwestern.edu
    affiliation:
      - name: Northwestern University
        city: Evanston
        state: IL
        url: https://www.northwestern.edu
date: 2024-06-03
date-format: long
format: 
  html:
    toc: true
  pdf:
    linestretch: 1.8
    # indent: true
    output-file: "Feinleib_senate_coalitions.pdf"
keywords: 
  - U.S. Senate
  - legislative coalitions
  - filibuster
  - cloture
  - agent-based modeling
bibliography: abm_bib.json
csl: apa.csl
number-sections: true
number-depth: 3
execute:
  echo: false
citation: true
---

```{r packages}
#| message: false

## packages used
library(knitr)
library(rlang)
```


## Introduction

* Describe phenomenon
* Motivate
* Fit with ABM

## Background

The U.S. Congress as a whole, and the Senate more specifically, are unsurprisingly the subject of numerous formal models in political science.
A foundational model in this area is pivotal politics theory [@krehbielPivotalPoliticsTheory1998]. In pivotal politics theory, legislators get utility from bills based on how bills move policy toward their *ideal point* policy preferences. In my model, I use DW-NOMINATE scores from Voteview [@lewisVoteviewCongressionalRollcall2024] to measure these ideal points. In order for a bill to pass, the bill must move policy towards the ideal points of a majority of senators (and a three-fifths supermajority if a bill is filibustered). Pivotal politics and its conception of policy benefits are fundamental components of my agent-based model.

One of the defining features of the modern Senate is the filibuster and the cloture rule. Over time, the Senate's longstanding norms respecting senators' right to debate have evolved into the filibuster as we know it today: a tool of obstruction used by united minority parties against nearly all bills in the Senate [@jentlesonKillSwitchRise2022]. The filibuster is now so strong that it has created a "60-vote Senate," in which almost every bill requires a supermajority to pass [@kogerFilibusteringPoliticalHistory2010].
@wawroFilibusterObstructionLawmaking2006 explained the impact of the Senate's original 1917 cloture rule by modeling the incentives of legislative entrepreneurs before and after the introduction of the rule. This model is the foundation of my equation-based model [@feinleibRedesigningFilibusterMore2024]. Both of these models are explained further in @sec-ebm.
@kogerFilibusteringPoliticalHistory2010 responded to Wawro and Schickler by focusing more heavily on the costs of legislative obstruction (and the costs to the majority of withstanding it). Koger also presented a different model of the filibuster that is more explicitly based on game theory. My model incorporates Koger’s focus on costs into Wawro and Schickler’s original model.

Another theoretical building-block of my model is the role of cosponsorship in the lawmaking process.  Cosponsorship signals a strong commitment to support a bill: senators who cosponsor a bill almost never vote against it [@bernhardCommitmentConsequencesReneging2013]. @harbridge-yongBipartisanPathEffective2023 use cosponsor relationships to measure the level of bipartisanship in Congress. They introduce two measures of senators' bipartisanship that are included in my model:

* *Percent bipartisan cosponsors attracted (PBCA):* the average percentage of cosponsors that come from the opposite party on bills sponsored by a senator.
* *Percent bipartisan cosponsorships offered (PBCO):* the percentage of a senator's cosponsorships that go to bills sponsored by members of the opposite party.

Finally, my model aims to address a major outstanding puzzle in models on Congress. Many formal models, especially those based on pivotal politics, predict that bills will tend to pass with *minimal majorities*, that is, the smallest amount of support required by the rules of the Senate and House (three-fifths and one-half, respectively). The equation-based model in my thesis is one such model. However, these minimal majorities are relatively rare, empirically [@grosecloseBuyingSupermajorities1996; @uslanerPartisanshipCoalitionFormation1975]. Relatedly, even as the parties have become more polarized over the past few decades [@desilverPolarizationTodayCongress2022], the level of bipartisanship in bills that pass Congress has remained steady [@curryNonpartyGovernmentBipartisan2019].
An agent-based representation of my model, due to its heterogeneous treatment of individual members, has the potential to solve these issues with equation-based pivotal politics models.

## The Underlying Model {#sec-ebm}
The Senate Coalitions model is an agent-based representation of the equation-based model of legislative coalitions by @feinleibRedesigningFilibusterMore2024, which is in turn an expansion of Wawro and Schickler's [-@wawroFilibusterObstructionLawmaking2006] model of legislative entrepreneurship under cloture. Thus, understanding these models is essential for understanding many elements of the agent-based model. I will now proceed to describe those models, naturally borrowing heavily from the original papers.

As emphasized by Koger [-@kogerFilibusteringPoliticalHistory2010], the key factors in senators' decision-making during legislative fights are the relative costs and benefits of the different strategies available to them in such fights. Thus, to predict the outcomes of legislative battles, I compare the relative utilities of bill proponents and obstructionists under various rules environments.

### The Benefits for Proponents
The first element of my equation-based model is the benefits for bill proponents. This component comes directly from Wawro and Schickler [-@wawroFilibusterObstructionLawmaking2006], who model the expected benefits of bill proponents under the Senate cloture rule. This model, which I refer to as WS~K~,^[$K$ stands for cloture because $C$ refers to costs in my model. Wawro and Schickler also present a model without a cloture rule (labeled WS in @feinleibRedesigningFilibusterMore2024). I am skipping that model for brevity.] focuses on a legislative entrepreneur (LE) who receives benefits from passing their bills. The LE may add legislators to a coalition supporting their bill, which increases the probability that their proposal passes. However, growing the coalition decreases the benefits of passing the bill, as the LE may have to make compromises or share credit to gain additional support. Thus, legislative entrepreneurs face a trade-off when forming their proponent coalition.

The WS~K~ model is based on the following fundamentals:

* $\eta$: The size of the supporting coalition, as a proportion. 
  * $\eta$ is restricted to $\eta \in [.5, 1]$, as any bill must have at least majority support to pass. ^[For simplicity, WS~K~ ignores the case of a tie. In my agent-based model, I require bills to have strictly more than 50 votes to pass.]
  
* $\pi$: The probability the bill passes, which is a function of $\eta$:
$$
\pi(\alpha) = \left( \frac{\eta - .5}{.5} \right) ^ \alpha
$$

  * $\alpha$ is a parameter indicating how much each additional coalition member contributes to the probability of passage. In effect, this parameter controls the steepness of the $\pi$ curve. A higher value of $\alpha$ means that it takes more legislators in the supporting coalition to reach the same probability of passing a bill.
    * $\alpha \in (0, 1]$, so that as $\alpha$ increases toward 1, each additional legislator in the supporting coalition has a larger impact on the probability of passage.
  * $\pi$ is strictly increasing in $\eta$. Note that when $\eta = .50$, $\pi = 0$ (i.e., you need at least a minimum majority to pass a bill), and when $\eta = 1$, $\pi = 1$ (i.e., a bill with unanimous support is guaranteed to pass).
  
* $B$: The benefits to the LE of passing their bill, which is also a function of $\eta$. The benefits to the LE diminish as they add more supporters, as the LE may have to make compromises on the policy content of the bill or share credit with a larger group of supporters.
$$
B = \frac{1 - \eta}{.5}
$$
  * That is, $B$ linearly decreases from 1 to 0 as $\eta$ increases over $(0.5, 1]$.
  
  * Also, WS~K~ sets $B = 0$ if the proposal doesn't pass.
  
* $EB$: The expected benefits for the LE, which is simply the probability of passage multiplied by the benefits of passage:
$$
EB = \pi B
$$

Since $\pi$ is an increasing function in $\eta$, and $B$ is a decreasing function in $\eta$, there is an inherent tradeoff in forming an optimally sized coalition.

Finally, WS~K~ represents the Senate cloture rule as an increase in the passage probability when the supporting coalition clears the cloture threshold, $K$ (which is 0.60 under the Senate’s current rules).
When $\eta$ crosses $K$, the probability function switches from $\pi(\alpha)$ to $\pi(\alpha^*)$, with $\alpha^* < \alpha$, which scales up the probability.

Under WS~K~, the expected benefits are:

$$
EB_K = 
\begin{cases}
\pi(\alpha)B & \text{if } \eta < K \\
\pi(\alpha^*)B & \text{if } \eta ≥ K
\end{cases}
$$

@fig-ws-comp below illustrates how the expected benefits to the LE vary with the size of the supporting coalition. The plots on the left show a Senate without a cloture rule, and those on the right show a Senate with a cloture rule. @wawroFilibusterObstructionLawmaking2006 [p. 217] found that an $\alpha$ around 0.25 best fit the Senate of the late-19th and early-20th centuries, when the original cloture rule was introduced.
The plot shows how a cloture rule created a sharp increase in benefits at the 60-vote threshold. Thus, the WS~K~ model predicts that legislative entrepreneurs will seek supporting coalitions just large enough to clear the cloture threshold.

```{r WS helper functions}
prob_pass <- function(x, alpha) ((x - 0.5) / 0.5) ^ alpha
benefits <- function(x) (1 - x) / 0.5
eb_pass <- function(x, alpha) prob_pass(x, alpha) * benefits(x)

## plotting helpers
# linetypes legend
plot_legend <- function(x, y, linetypes, cex = 1) {
  legend(
    x = x, y = y, 
    legend = c(expr(pi), "B", 
               # EB or EB_K
               expr(!!parse_expr(gsub("_K", "[K]", names(linetypes)[3])))),
    lty = c(linetypes$pi, linetypes$B, linetypes$EB),
    box.col = "white",
    cex = cex
  )
}

# plot frame (empty plot)
plot_frame <- function(main, label_cex = 1) {
  plot(NULL, xlim = c(0.51, 1), ylim = c(0,1),
       xlab = "Coalition size", 
       ylab = "Probability of passage/expected benefits",
       main = main,
       # control size of labels
       cex.axis = label_cex,
       cex.lab = label_cex,
       cex.main = label_cex)
  
  # control size of axis tick labels
  axis(1, cex.axis = label_cex)
  axis(2, cex.axis = label_cex)
}

# plot curves
plot_curves <- function(pi_fn, b_fn, eb_fn, linetypes, ...) {
  curve(lapply(x, \(x) pi_fn(x, ...)), 
        add = T, lty = linetypes$pi)
  curve(b_fn,
        add = T, lty = linetypes$B)
  curve(lapply(x, \(x) eb_fn(x, ...)),
        add = T, lty = linetypes$EB)
}

# plot max EB
plot_max_eb <- function(eb_fn, label_cex = 1, ...) {
  max_eb <- which.max(vapply(X = seq(0.51, 1, 0.01), 
                             FUN = \(x) eb_fn(x, ...),
                             FUN.VALUE = numeric(1))) * 0.01 + 0.5
  points(max_eb, eb_fn(max_eb, ...), 
         # diamond shape
         pch = 18, 
         cex = 2 * label_cex)
}
```

```{r WS_K helper functions}
prob_pass_k <- function(x, alpha, alpha_star, k) {
  ifelse(x < k,
         prob_pass(x, alpha),
         prob_pass(x, alpha_star))
}

eb_pass_k <- function(x, alpha, alpha_star, k) {
  prob_pass_k(x, alpha, alpha_star, k) * benefits(x)
}
```

```{r main plotting functions}
plot_eb <- function(alpha,
                    leg_x = 0.85, leg_y = 0.75, 
                    pdf_cex = 0.7, html_cex = 1,
                    pdf_label_cex = 0.8, html_label_cex = 1) {
  linetypes <- list(pi = "dashed",
                    B = "dotted",
                    EB = "solid")
  
  label_cex <- ifelse(knitr::is_latex_output(),
                      pdf_label_cex,
                      html_label_cex)
  legend_cex <- ifelse(knitr::is_latex_output(),
                       pdf_cex,
                       html_cex)
  
  # plot frame
  plot_frame(main = expr(paste(alpha, " = ", !!alpha)),
             label_cex = label_cex)
  # plot curves (pi, B, EB)
  plot_curves(pi_fn = prob_pass, b_fn = benefits, eb_fn = eb_pass, 
              linetypes = linetypes, alpha = alpha)
  # plot linetype legend
  plot_legend(x = leg_x, y = leg_y, linetypes = linetypes,
              # control legend size depending on output format
              cex = legend_cex)
  # plot max EB
  plot_max_eb(eb_pass, alpha = alpha, label_cex = label_cex)
}

plot_eb_k <- function(alpha, alpha_star, k,
                      leg_x = 0.8, leg_y = 0.75, 
                      pdf_cex = 0.7, html_cex = 1,
                      pdf_label_cex = 0.8, html_label_cex = 1) {
  linetypes <- list(pi = "dashed",
                    B = "dotted",
                    EB_K = "solid")
  
  label_cex <- ifelse(knitr::is_latex_output(),
                      pdf_label_cex,
                      html_label_cex)
  legend_cex <- ifelse(knitr::is_latex_output(),
                       pdf_cex,
                       html_cex)
  
  # plot frame
  plot_frame(main = expr(paste(alpha, " = ", !!alpha, ", ",
                               alpha, "*", " = ", !!alpha_star)),
             label_cex = label_cex)
  # plot curves (pi, B, EB_K)
  plot_curves(pi_fn = prob_pass_k, b_fn = benefits, eb_fn = eb_pass_k, 
              linetypes = linetypes, alpha = alpha, 
              k = k, alpha_star = alpha_star)
  # plot linetype legend
  plot_legend(x = leg_x, y = leg_y, linetypes = linetypes,
              # control legend size depending on output format
              cex = legend_cex)
  # plot max EB
  plot_max_eb(eb_pass_k, 
              alpha = alpha, alpha_star = alpha_star, k = k,
              label_cex = label_cex)
}
```

```{r plot WS vs WS_K}
#| label: fig-ws-comp
#| fig-cap: "Proponents' expected benefits without (left) and with (right) a cloture rule"
#| fig-asp: 1.1

par(mfrow = c(2,2))

plot_eb(0.1)
plot_eb_k(alpha = 0.1, alpha_star = 0.05, k = 0.60)
plot_eb(alpha = 0.25)
plot_eb_k(alpha = 0.25, alpha_star = 0.1, k = 0.60)
```

### The Benefits for Obstructionists
In a legislative debate, the opponents are the remaining senators who do not support a bill. However, not all these opponents are willing to engage in active obstruction to prevent a bill from coming to a vote. 
These obstructionists are a subset of the bill opponents. As the opponents are a group with size $1 − \eta$, I label the size of the obstructionist faction $\omega \in [0, 1 − \eta]$.

Similar to bill proponents, the obstructionists receive benefits if they successfully kill a bill. These blocking benefits represent the obstructionists' preference for the status quo over the new policy proposal (the opposite of the proponents' policy benefits). They also may receive position-taking benefits for the act of obstruction regardless of the bill's outcome. By committing effort to actively obstruct a bill, obstructionists publicly demonstrate the strength of their policy commitments, which may improve their reputation with voters [@gibbsReputationPoliticsFilibuster2023]. Thus, the expected benefits of obstruction are:
$$
EB_{obst} = (1 - \pi)\beta_{block} + \beta_{position}
$$
where $\beta_{block}$ and $\beta_{position}$ represent the obstructionists' benefits for blocking a bill and their position-taking benefits, respectively. These terms are measured relative to the proponents' benefits $B$, which are fixed to a maximum of 1.

As with proponents, the probability and benefits of success for obstructionists depend on the size of the obstructionist faction.
Both forms of benefits for the obstructionists are shared among all the members of the obstructionist faction. Thus, the benefits for each individual obstructionist are:
$$
\beta_{block} = \left( \frac{.5 - \omega}{.5} \right) \sum\beta_{block}
\qquad
\beta_{position} = \left( \frac{.5 - \omega}{.5} \right) \sum\beta_{position}
$$
where $\sum\beta_{block}$ and $\sum\beta_{position}$ are the total benefits of obstruction shared among the obstructionists.
For each $\beta$, if there is a single obstructionist, then that member alone accrues the entire benefit. If $\omega = 0.50$ (i.e., the obstructionists make up half the chamber), then obstruction produces no benefits, as in that case, the bill opponents already have enough votes to block a bill without any additional obstruction.

### The Costs for Proponents and Obstructionists
In order to provide a full accounting of the expected utility for bill proponents and obstructionists, we must also consider the costs each side faces.
According to Koger's [-@kogerFilibusteringPoliticalHistory2010] analysis of Senate filibustering, the primary factor in these costs is time. Time represents a cost to senators in two forms: the consumption of the finite floor time available during a legislative session and the opportunity costs related to other activities senators could be doing, such as campaigning and fundraising.

The costs associated with passing or obstructing a bill depend heavily on the rules and procedures of the Senate. As the Senate's rules surrounding filibustering and cloture have evolved, so too has the amount of time it takes to debate and vote on a bill.
In a legislative fight, each faction $i$ faces a time cost $C_i$ of engaging in the fight. The costs to members of the proponent, opponent, and obstructionist factions are denoted as $C_{prop}$, $C_{opp}$, and $C_{obst}$, respectively.
The time cost $C_i$ of considering a bill can be broken down into the following components, such that $C_i = C_{i,t} + C_{i,d} + C_{i,v}$:

* $C_{i,t}$: The cost of the consumption of floor time.
* $C_{i,d}$: The cost of physical attendance and debate time.
* $C_{i,v}$: The cost of voting time.

These costs may be different for each faction. For example, in a talking filibuster, the obstructionists face heavy debate costs from having to continuously speak, while the proponents and opponents just have to wait around (i.e., $C_{obst, d} > C_{prop, d}$ and $C_{obst, d} > C_{opp, d}$).
Different rules around filibustering and cloture would change the amount of time the Senate spends on each bill. Given a specific set of rules, the amount of time spent on a bill and its cost also depend on the sizes of the proponent and obstructionist factions.

For demonstration, consider the costs of considering bills under the current rules with a few different mixes of proponents and obstructionists:

1. All senators support a bill ($\eta = 1$). The proponents can pass the bill by unanimous consent and spend no floor time, resulting in $C = 0$ for all factions.
2. A non-unanimous majority of senators supports a bill, but none of the opponents are willing to actively obstruct it ($\eta \in (0.5, 1), \ \omega = 0$). In order to pass the bill, the proponents must spend some floor time for debate and a single passage vote. Since there are no obstructionists, the opponents are submitting to a passage vote without forcing a cloture vote. This results in a small $C$ for the proponents.
3. A filibuster-proof majority of senators supports a bill, and at least one senator obstructs a bill ($\eta \in [0.6, 1), \ \omega > 0$). In this case, proponents may either spend floor time on holding a cloture vote, or wait for the obstructionist to debate until they submit to a final passage vote. The second option is rare in the modern Senate because of the high cost in floor time, but it is still an available strategy [@kogerFilibusteringPoliticalHistory2010]. For the obstructionists, debate would be even more costly, although they face no additional costs in the case of a cloture vote.
4. The supporting coalition is smaller than the cloture threshold, and at least one senator obstructs a bill ($\eta \in (0.5, 0.6), \ \omega > 0$). Proponents would fail a cloture vote, so their only option for passing a bill is waiting out the obstructionists. Again, this is extremely costly in floor time ($C_{prop, t}$). Accordingly, senators rarely employ this strategy in the modern Senate. This is why it is nearly impossible to pass a bill with fewer than 60 supporters in today's climate.

Different rules environments would allow for different strategies and different costs.

### Comparing the Expected Utility of the Two Factions
To predict the outcome of legislative battles, I compare the total expected utility of the members of the proponent and obstructionist factions. Since each member of the proponent and obstructionist factions receives expected benefits $EB_i$ and costs $C_i$, the following expressions give the aggregate expected utility for each faction:

$$
\begin{aligned}
\sum EU_{prop} &= \eta   \cdot (EB_{prop} - C_{prop}) \\
\sum EU_{obst} &= \omega \cdot (EB_{obst} - C_{obst})
\end{aligned}
$$
In this model, it is important to sum the benefits and costs among the members of each faction because those benefits and costs are shared. As a faction grows, even though an individual senator's benefits and costs decrease, the total benefits and costs increase with the size of a faction.

If each faction receives expected utility $\sum EU_i$ from playing their respective strategy (that is, by representing a proponent or obstructionist faction), then the side with the greater (positive) expected utility will be more willing to engage in the legislative fight.^[If a faction receives negative expected utility, then they would simply surrender the fight.]
This utility-based method to predict the winner of a legislative fight is possible because senators have the ability to ensure a win at great cost (proponents by changing the rules or forcing the obstructionists to continuously debate, and obstructionists by being willing to continuously debate) if they evaluate that winning the fight produces enough benefits.

At first glance, this rule to predict winners using expected utility may appear rigid, as if either the proponents or obstructionists are doomed from the start of a legislative fight. How does this rule explain the negotiation process that is so critical for proponents to grow the coalition to pass a bill, or for obstructionists to kill it? The answer is that the expected utility of these two factions constantly varies as the factions grow and shrink and they change their cost-benefit calculations. As the coalitions and the political environment change, and more and more time passes, the expected-utility winner may shift. This change over time is what makes the expected-utility system dynamic.

## An Agent-Based Representation

The Senate Coalitions ABM is a agent-based representation of the equation-based model described above. While the agent-based model does not fully reproduce all the elements of the equation-based model, its behavior resembles the expected behavior of the equation-based model and that of the real-life Senate. I will now explain the rules of the Senate Coalitions model.

### Setup
The most substantive portion of the `SETUP` procedure is the initialization of senators. Senators are created using data from @harbridge-yongBipartisanPathEffective2023 on the 114th Congress (2015--16). The following senator properties come from this data: first name, last name, home state, DW-NOMINATE 1st-dimension estimate, DW-NOMINATE 2nd-dimension estimate, party, average number of cosponsors on sponsored bills, PBCA, and PBCO.
Senators' locations in the view are based on their DW-NOMINATE coordinates, and they are colored by party (blue for Democrats, red for Republicans, and yellow for independents). Senators from the most common party are labeled as majority-party senators. All senators are placed in an empty coalition.
The `SETUP` procedure also clears previous model state, resets the tick counter, and resets the metrics that are used in the monitors and plots in the interface.

### Go
The `GO` procedure runs through the introduction and consideration of a bill. Unlike in most NetLogo models, the `GO` procedure does not last just one tick. Instead, each of the below sub-procedures, each of which represents one step in the legislative process, increments the tick counter. This design gives the model a richer conception of the time it takes to consider a bill.

#### Creating a Bill
The `PLACE-BILL` procedure creates a new bill. First, it creates a status quo policy at a random point in DW-NOMINATE policy space. Then it creates a bill at another random point in policy space. Thus, the bills that are introduced are totally random and independent of the policy preferences of all senators. Bills that make little political sense are introduced, but these reliably fail.

The new bill is initialized with no sponsor or cosponsors. The new bill adopts the just-created status quo as its `squo-policy`, and creates an arrow-shaped link from that status quo. The new bill becomes the active bill.

The tick counter is incremented, and the `active-time` of the bill is set to 1.

#### Attracting a Sponsor and Cosponsors
This step occurs in the `GET-SPONSOR-AND-COSPONSORS` procedure.

##### Sponsor
All bills in the Senate must be introduced by a sponsoring senator. It is reasonable to assume that senators introduce bills that align with their policy preferences. Thus, the bill selects the senator with the greatest benefits from its policy movement to be the sponsor. The bill creates a green link with its sponsor, and the sponsor permanently becomes a proponent of this bill.

##### Cosponsors
The bill then proceeds to attract cosponsors. The cosponsorship-attraction process is based on the interaction of the bill's policy movement, the properties of the bill's sponsor, and the properties of the other 99 senators. First, the number of cosponsors (`n-cosponsors`) is randomly drawn from a Poisson distribution using the sponsor’s average number of cosponsors on their bills.

Next, the 99 other senators calculate their "cosponsor likelihood" for the bill based on three factors.
First, the senator calculates their policy benefits from the bill.
Then, the senator multiplies those benefits by a "party factor" based on the interaction of their own party and the sponsor's party. The party factor is the product of the senator's frequency of offering cosponsorships to members of the sponsor's party (PBCO) and the sponsor's frequency of attracting cosponsorships from member's of the senator's party (PBCA). If either the senator or the sponsor is an Independent, both the PBCO and PBCA are treated as 50%. This calculation considers the varying behaviors of senators related to bipartisanship.
Finally, a small bonus is added if the senator is from the same home state as the sponsor. It is common for senators from the same state to work together on legislation, and this bonus represents that.

Note that this "cosponsor likelihood" is not a proper 0--1 probability. It is a real number that may be positive or negative, and take on quite large magnitudes. These facts are not significant, though, as the cosponsor likelihood is simply used for ordering.

The bill sets its cosponsors agentset to be the `n-cosponsors` senators with the highest positive cosponsor likelihoods. If less than N-COSPONSORS senators have positive cosponsor likelihoods, the cosponsors will just be those senators with positive cosponsor likelihoods.
The cosponsors create gray links with the sponsor, and the cosponsors permanently become proponents of the bill. This commitment to the bill is based on the reliable support of cosponsors in the real-life Congress [@bernhardCommitmentConsequencesReneging2013].

The tick counter and the `active-time` of the bill are both incremented.

#### Creating Initial Coalitions


#### Iterating Coalitions

Benefits:

* heterogenous treatment of members

Simplifications/omissions:

* looking at individual members' utilities, not the total expected utility of the coalitions
* Not considering the various strategies available to senators/coalitions

Sections:

* Rules of the model
* Justifications for rules
* Model parameters
* Simplifications and assumption

## Results

* Experiments you ran (parameters, # of repetitions)
* Results from those experiments (include graphs)
* Empirical validation (if applicable)

## Analysis and Discussion

* Interpretation of results (include graphs)
* What emergent phenomena arose? (include screenshots)
* Implications

## Conclusion

* What would future work on this look like?
* What might you suggest for further validation?

## Software Credits {.appendix}

This agent-based model was built in the NetLogo programming language [@wilenskyNetLogo1999].
The model uses the NetLogo CSV extension [@CSVExtension2024], and borrows code for that extension from the CSV Example model in the NetLogo Models Library [@wilenskyCSVExample].
Data used in the model comes from @harbridge-yongBipartisanPathEffective2023 and the Legislative Effectiveness Scores project [@voldenLegislativeEffectivenessScores2023]. This data was retrieved using the `filibustr` R package [@feinleibFilibustrDataUtilities2024].

\pagebreak

## References {.appendix}
