---
title: Senate Coalitions
subtitle: An Agent-Based Model
author: 
  - name: 
      given: Max H. 
      family: Feinleib
    orcid: 0009-0002-9604-3533
    email: maxfeinleib2024@u.northwestern.edu
    affiliation:
      - name: Northwestern University
        city: Evanston
        state: IL
        url: https://www.northwestern.edu
date: 2024-06-03
date-format: long
format: 
  html: default
  pdf:
    linestretch: 1.8
    # indent: true
    output-file: "Feinleib_senate_coalitions.pdf"
keywords: 
  - U.S. Senate
  - legislative coalitions
  - filibuster
  - cloture
  - agent-based modeling
bibliography: abm_bib.json
csl: apa.csl
toc: true
number-sections: true
execute:
  echo: false
citation: true
---

\pagebreak

```{r packages}
#| message: false

## packages used
library(knitr)
library(rlang)
```


## Introduction

* Describe phenomenon
* Motivate
* Fit with ABM

## Background

The U.S. Congress as a whole, and the Senate more specifically, are unsurprisingly the subject of numerous formal models in political science.
A foundational model in this area is pivotal politics theory [@krehbielPivotalPoliticsTheory1998]. In pivotal politics theory, legislators get utility from bills based on how bills move policy toward their *ideal point* policy preferences. In my model, I use DW-NOMINATE scores from Voteview [@lewisVoteviewCongressionalRollcall2024] to measure these ideal points. In order for a bill to pass, the bill must move policy towards the ideal points of a majority of senators (and a three-fifths supermajority if a bill is filibustered). Pivotal politics and its conception of policy benefits are fundamental components of my agent-based model.

One of the defining features of the modern Senate is the filibuster and the cloture rule. Over time, the Senate's longstanding norms respecting senators' right to debate have evolved into the filibuster as we know it today: a tool of obstruction used by united minority parties against nearly all bills in the Senate [@jentlesonKillSwitchRise2022]. The filibuster is now so strong that it has created a "60-vote Senate," in which almost every bill requires a supermajority to pass [@kogerFilibusteringPoliticalHistory2010].
@wawroFilibusterObstructionLawmaking2006 explained the impact of the Senate's original 1917 cloture rule by modeling the incentives of legislative entrepreneurs before and after the introduction of the rule. This model is the foundation of my equation-based model [@feinleibRedesigningFilibusterMore2024]. Both of these models are explained further in @sec-ebm.
@kogerFilibusteringPoliticalHistory2010 responded to Wawro and Schickler by focusing more heavily on the costs of legislative obstruction (and the costs to the majority of withstanding it). Koger also presented a different model of the filibuster that is more explicitly based on game theory. My model incorporates Koger’s focus on costs into Wawro and Schickler’s original model.

Another theoretical building-block of my model is the role of cosponsorship in the lawmaking process.  Cosponsorship signals a strong commitment to support a bill: senators who cosponsor a bill almost never vote against it [@bernhardCommitmentConsequencesReneging2013]. @harbridge-yongBipartisanPathEffective2023 use cosponsor relationships to measure the level of bipartisanship in Congress. They introduce two measures of senators' bipartisanship that are included in my model:

* *Percent bipartisan cosponsors attracted (PBCA):* the average percentage of cosponsors that come from the opposite party on bills sponsored by a senator.
* *Percent bipartisan cosponsorships offered (PBCO):* the percentage of a senator's cosponsorships that go to bills sponsored by members of the opposite party.

Finally, my model aims to address a major outstanding puzzle in models on Congress. Many formal models, especially those based on pivotal politics, predict that bills will tend to pass with *minimal majorities*, that is, the smallest amount of support required by the rules of the Senate and House (three-fifths and one-half, respectively). The equation-based model in my thesis is one such model. However, these minimal majorities are relatively rare, empirically [@grosecloseBuyingSupermajorities1996; @uslanerPartisanshipCoalitionFormation1975]. Relatedly, even as the parties have become more polarized over the past few decades [@desilverPolarizationTodayCongress2022], the level of bipartisanship in bills that pass Congress has remained steady [@curryNonpartyGovernmentBipartisan2019].
An agent-based representation of my model, due to its heterogeneous treatment of individual members, has the potential to solve these issues with equation-based pivotal politics models.

## The Underlying Model {#sec-ebm}
The Senate Coalitions model is an agent-based representation of the equation-based model of legislative coalitions by @feinleibRedesigningFilibusterMore2024, which is in turn an expansion of Wawro and Schickler's [-@wawroFilibusterObstructionLawmaking2006] model of legislative entrepreneurship under cloture. Thus, understanding these models is essential for understanding many elements of the agent-based model. I will now proceed to describe those models, naturally borrowing heavily from the original papers.

As emphasized by Koger [-@kogerFilibusteringPoliticalHistory2010], the key factors in senators' decision-making during legislative fights are the relative costs and benefits of the different strategies available to them in such fights. Thus, to predict the outcomes of legislative battles, I compare the relative utilities of bill proponents and obstructionists under various rules environments.

### The Benefits for Proponents
Wawro and Schickler [-@wawroFilibusterObstructionLawmaking2006] model the expected benefits of bill proponents under the Senate cloture rule. This model, which I refer to as WS~K~,^[$K$ stands for cloture because $C$ refers to costs in my model. Wawro and Schickler also present a model without a cloture rule (labeled WS in @feinleibRedesigningFilibusterMore2024); I am skipping that model for brevity.] focuses on a legislative entrepreneur (LE) who receives benefits from passing their bills. The LE may add legislators to a coalition supporting their bill, which increases the probability that their proposal passes. However, growing the coalition decreases the benefits of passing the bill, as the LE may have to make compromises or share credit to gain additional support. Thus, legislative entrepreneurs face a trade-off when forming their proponent coalition.

The WS model is based on the following fundamentals:

* $\eta$: The size of the supporting coalition, as a proportion. 
  * $\eta$ is restricted to $\eta \in [.5, 1]$, as any bill must have at least majority support to pass. ^[For simplicity, WS ignores the case of a tie. In my agent-based model, I require bills to have strictly more than 50 votes to pass.]
  
* $\pi$: The probability the bill passes, which is a function of $\eta$:
$$
\pi(\alpha) = \left( \frac{\eta - .5}{.5} \right) ^ \alpha
$$

  * $\alpha$ is a parameter indicating how much each additional coalition member contributes to the probability of passage. In effect, this parameter controls the steepness of the $\pi$ curve. A higher value of $\alpha$ means that it takes more legislators in the supporting coalition to reach the same probability of passing a bill.
    * $\alpha \in (0, 1]$, so that as $\alpha$ increases toward 1, each additional legislator in the supporting coalition has a larger impact on the probability of passage.
  * $\pi$ is strictly increasing in $\eta$. Note that when $\eta = .50$, $\pi = 0$ (i.e., you need at least a minimum majority to pass a bill), and when $\eta = 1$, $\pi = 1$ (i.e., a bill with unanimous support is guaranteed to pass).
  
* $B$: The benefits to the LE of passing their bill, which is also a function of $\eta$. The benefits to the LE diminish as they add more supporters, as the LE may have to make compromises on the policy content of the bill or share credit with a larger group of supporters.
$$
B = \frac{1 - \eta}{.5}
$$
  * That is, $B$ linearly decreases from 1 to 0 as $\eta$ increases over $(0.5, 1]$.
  
  * Also, WS~K~ sets $B = 0$ if the proposal doesn't pass.
  
* $EB$: The expected benefits for the LE, which is simply the probability of passage multiplied by the benefits of passage:
$$
EB = \pi B
$$

Since $\pi$ is an increasing function in $\eta$, and $B$ is a decreasing function in $\eta$, there is an inherent tradeoff in forming an optimally sized coalition.

Finally, WS~K~ represents the Senate cloture rule as an increase in the passage probability when the supporting coalition clears the cloture threshold, $K$ (which is 0.60 under the Senate’s current rules).
When $\eta$ crosses $K$, the probability function switches from $\pi(\alpha)$ to $\pi(\alpha^*)$, with $\alpha^* < \alpha$, which scales up the probability.

Under WS~K~, the expected benefits are:

$$
EB_K = 
\begin{cases}
\pi(\alpha)B & \text{if } \eta < K \\
\pi(\alpha^*)B & \text{if } \eta ≥ K
\end{cases}
$$

@fig-ws-comp below illustrates how the expected benefits to the LE vary with the size of the supporting coalition. The plots on the left show a Senate without a cloture rule, and those on the right show a Senate with a cloture rule. @wawroFilibusterObstructionLawmaking2006 [p. 217] found that an $\alpha$ around 0.25 best fit the Senate of the late-19th and early-20th centuries, when the original cloture rule was introduced.
The plot shows how a cloture rule created a sharp increase in benefits at the 60-vote threshold. Thus, the WS~K~ model predicts that legislative entrepreneurs will seek supporting coalitions just large enough to clear the cloture threshold.

```{r WS helper functions}
prob_pass <- function(x, alpha) ((x - 0.5) / 0.5) ^ alpha
benefits <- function(x) (1 - x) / 0.5
eb_pass <- function(x, alpha) prob_pass(x, alpha) * benefits(x)

## plotting helpers
# linetypes legend
plot_legend <- function(x, y, linetypes, cex = 1) {
  legend(
    x = x, y = y, 
    legend = c(expr(pi), "B", 
               # EB or EB_K
               expr(!!parse_expr(gsub("_K", "[K]", names(linetypes)[3])))),
    lty = c(linetypes$pi, linetypes$B, linetypes$EB),
    box.col = "white",
    cex = cex
  )
}

# plot frame (empty plot)
plot_frame <- function(main, label_cex = 1) {
  plot(NULL, xlim = c(0.51, 1), ylim = c(0,1),
       xlab = "Coalition size", 
       ylab = "Probability of passage/expected benefits",
       main = main,
       # control size of labels
       cex.axis = label_cex,
       cex.lab = label_cex,
       cex.main = label_cex)
  
  # control size of axis tick labels
  axis(1, cex.axis = label_cex)
  axis(2, cex.axis = label_cex)
}

# plot curves
plot_curves <- function(pi_fn, b_fn, eb_fn, linetypes, ...) {
  curve(lapply(x, \(x) pi_fn(x, ...)), 
        add = T, lty = linetypes$pi)
  curve(b_fn,
        add = T, lty = linetypes$B)
  curve(lapply(x, \(x) eb_fn(x, ...)),
        add = T, lty = linetypes$EB)
}

# plot max EB
plot_max_eb <- function(eb_fn, label_cex = 1, ...) {
  max_eb <- which.max(vapply(X = seq(0.51, 1, 0.01), 
                             FUN = \(x) eb_fn(x, ...),
                             FUN.VALUE = numeric(1))) * 0.01 + 0.5
  points(max_eb, eb_fn(max_eb, ...), 
         # diamond shape
         pch = 18, 
         cex = 2 * label_cex)
}
```

```{r WS_K helper functions}
prob_pass_k <- function(x, alpha, alpha_star, k) {
  ifelse(x < k,
         prob_pass(x, alpha),
         prob_pass(x, alpha_star))
}

eb_pass_k <- function(x, alpha, alpha_star, k) {
  prob_pass_k(x, alpha, alpha_star, k) * benefits(x)
}
```

```{r main plotting functions}
plot_eb <- function(alpha,
                    leg_x = 0.85, leg_y = 0.75, 
                    pdf_cex = 0.7, html_cex = 1,
                    pdf_label_cex = 0.8, html_label_cex = 1) {
  linetypes <- list(pi = "dashed",
                    B = "dotted",
                    EB = "solid")
  
  label_cex <- ifelse(knitr::is_latex_output(),
                      pdf_label_cex,
                      html_label_cex)
  legend_cex <- ifelse(knitr::is_latex_output(),
                       pdf_cex,
                       html_cex)
  
  # plot frame
  plot_frame(main = expr(paste(alpha, " = ", !!alpha)),
             label_cex = label_cex)
  # plot curves (pi, B, EB)
  plot_curves(pi_fn = prob_pass, b_fn = benefits, eb_fn = eb_pass, 
              linetypes = linetypes, alpha = alpha)
  # plot linetype legend
  plot_legend(x = leg_x, y = leg_y, linetypes = linetypes,
              # control legend size depending on output format
              cex = legend_cex)
  # plot max EB
  plot_max_eb(eb_pass, alpha = alpha, label_cex = label_cex)
}

plot_eb_k <- function(alpha, alpha_star, k,
                      leg_x = 0.8, leg_y = 0.75, 
                      pdf_cex = 0.7, html_cex = 1,
                      pdf_label_cex = 0.8, html_label_cex = 1) {
  linetypes <- list(pi = "dashed",
                    B = "dotted",
                    EB_K = "solid")
  
  label_cex <- ifelse(knitr::is_latex_output(),
                      pdf_label_cex,
                      html_label_cex)
  legend_cex <- ifelse(knitr::is_latex_output(),
                       pdf_cex,
                       html_cex)
  
  # plot frame
  plot_frame(main = expr(paste(alpha, " = ", !!alpha, ", ",
                               alpha, "*", " = ", !!alpha_star)),
             label_cex = label_cex)
  # plot curves (pi, B, EB_K)
  plot_curves(pi_fn = prob_pass_k, b_fn = benefits, eb_fn = eb_pass_k, 
              linetypes = linetypes, alpha = alpha, 
              k = k, alpha_star = alpha_star)
  # plot linetype legend
  plot_legend(x = leg_x, y = leg_y, linetypes = linetypes,
              # control legend size depending on output format
              cex = legend_cex)
  # plot max EB
  plot_max_eb(eb_pass_k, 
              alpha = alpha, alpha_star = alpha_star, k = k,
              label_cex = label_cex)
}
```

```{r plot WS vs WS_K}
#| label: fig-ws-comp
#| fig-cap: "Proponents' expected benefits without (left) and with (right) a cloture rule"
#| fig-asp: 1.1

par(mfrow = c(2,2))

plot_eb(0.1)
plot_eb_k(alpha = 0.1, alpha_star = 0.05, k = 0.60)
plot_eb(alpha = 0.25)
plot_eb_k(alpha = 0.25, alpha_star = 0.1, k = 0.60)
```

### The Benefits for Obstructionists
In a legislative debate, the opponents are the remaining senators who do not support a bill. However, not all these opponents are willing to engage in active obstruction to prevent a bill from coming to a vote. 
These obstructionists are a subset of the bill opponents. As the opponents are a group with size $1 − \eta$, I label the size of the obstructionist faction $\omega \in [0, 1 − \eta]$.

Similar to bill proponents, the obstructionists receive benefits if they successfully kill a bill. These blocking benefits represent the obstructionists' preference for the status quo over the new policy proposal (the opposite of the proponents' policy benefits). They also may receive position-taking benefits for the act of obstruction regardless of the bill's outcome. By committing effort to actively obstruct a bill, obstructionists publicly demonstrate the strength of their policy commitments, which may improve their reputation with voters [@gibbsReputationPoliticsFilibuster2023]. Thus, the expected benefits of obstruction are:
$$
EB_{obst} = (1 - \pi)\beta_{block} + \beta_{position}
$$


## An Agent-Based Representation

* Rules of the model
* Justifications for rules
* Model parameters
* Simplifications and assumption

## Results

* Experiments you ran (parameters, # of repetitions)
* Results from those experiments (include graphs)
* Empirical validation (if applicable)

## Analysis and Discussion

* Interpretation of results (include graphs)
* What emergent phenomena arose? (include screenshots)
* Implications

## Conclusion

* What would future work on this look like?
* What might you suggest for further validation?

## Software Credits {.appendix}

This agent-based model was built in the NetLogo programming language [@wilenskyNetLogo1999].
The model uses the NetLogo CSV extension [@CSVExtension2024], and borrows code for that extension from the CSV Example model in the NetLogo Models Library [@wilenskyCSVExample].
Data used in the model comes from @harbridge-yongBipartisanPathEffective2023 and the Legislative Effectiveness Scores project [@voldenLegislativeEffectivenessScores2023]. This data was retrieved using the `filibustr` R package [@feinleibFilibustrDataUtilities2024].

\pagebreak

## References {.appendix}
